---
title: "Analyze Eagle Activity using Hidden Markov Models"
author: "Mukai Wang 98830336"
date: "4/23/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(pomp)
library(knitr)
library(ggplot2)
library(fitdistrplus)
setwd(getwd())
```

# Introduction

This final project is a replication of a previous research paper which took advantage of hidden markov models. The article's title is "Analysis of animal accelerometer data using hidden Markov Models". Using accelerometer to capture animals' motion is a prevalent way to keep track of animal behaviour. The measured data, usually called minimum specific acceleration(MSA), can directly indicate the instant activeness of the animal. The larger the MSA is, the probable that the animal is in an active state. I am going to use one of the dataset from this research article to experiment hidden Markov models and sketch out the activeness of an Verreaux eagle throughout a day.

# Data and Model Background

According to the article, an Verreaux eagle(simplified as "eagle" below) was tracked for 9 days from 04/16 to 04/24. On each day data were collected from about 8am to 8pm, with a time interval of about one and a half minutes between two adjacent timepoints. Therefore each time series from different days are independent of each other.

According to the article, the authors presume that there are two hidden states (active and inactive). Within the active state, the MSA follows a gamma distribution. Within the inactive state, the MSA follows a mixture of two gamma distributions. Gamma distribution is a simple and ideal disribution to describe a set of points that are mostly small with a handful of extremely large points. The authors decide to use a mixture of gamma distributions for the inactive state because the eagle could be either gliding at a low speed or it could be roosting.

Since there are only two states, the state transition can be represented by a 2 by 2 probability matrix. In the dataset the authors also collected the wind speed and temperature for each time point. They believed that the transitional probability can be fitted by a logistic regression with wind speed, temperature and their interaction.

# Analysis Scheme

## Model Assumption
Because of the time constraint, I decide to carry out analysis only on data points collect on a single day. Since there are the most number of data points on 04/21 (235). My plan is to fit seven POMP models, compare their likelihood, then interpret the one with the best fit.

First I plan to just fit a gamma distribution on the 235 data points.

The second model I want to fit is the simplest POMP model. It has a transitional probability matrix which is a constant.
$$
\begin{bmatrix}
1-p_{0} & p_{0}\\
1-p_{1} & p_{1}
\end{bmatrix}
$$
$p_{0}$ represents the probability of the eagle changing from inactive state to active state $p_{1}$ represents the probability of the eagle staying in the active state. Within each state the MSA follows its own gamma distribution
$$
\begin{aligned}
\text{MSA} \stackrel{\text{inactive}}{\sim} Gamma(\text{shape0}, \text{scale0})\\
\text{MSA} \stackrel{\text{active}}{\sim} Gamma(\text{shape1}, \text{scale1})
\end{aligned}
$$

The third model is similar, but differs from the second model in that the measurement distribution of the inactive state is a mixture of two gamma distributions

$$
\text{MSA} \stackrel{\text{inactive}}{\sim} p_{mix}\times Gamma(\text{shape01}, \text{scale01})+(1-p_{mix})\times Gamma(\text{shape02}, \text{scale02})
$$

The fourth model adds to the third model by making the transitional probability follow a logistic regression with wind speed as a covariate, namely

$$
\begin{aligned}
\text{logit}(p_{0})\sim \beta_{00}+\beta_{01}\text{wind}\\
\text{logit}(p_{1})\sim \beta_{10}+\beta_{11}\text{wind}
\end{aligned}
$$

The fifth model has the logistic regression based on temperature

$$
\begin{aligned}
\text{logit}(p_{0})\sim \beta_{00}+\beta_{02}\text{temperature}\\
\text{logit}(p_{1})\sim \beta_{10}+\beta_{12}\text{temperature}
\end{aligned}
$$

The sixth model has the logistic regression based on the main effects of both wind speed and temperature

$$
\begin{aligned}
\text{logit}(p_{0})\sim \beta_{00}+\beta_{01}\text{wind}+\beta_{02}\text{temperature}\\
\text{logit}(p_{1})\sim \beta_{10}+\beta_{11}\text{wind}+\beta_{12}\text{temperature}
\end{aligned}
$$

The seventh model is the full model with both covariates and their interaction term

$$
\begin{aligned}
\text{logit}(p_{0})\sim \beta_{00}+\beta_{01}\text{wind}+\beta_{02}\text{temperature}+\beta_{03}\text{temperature}\times \text{wind}\\
\text{logit}(p_{1})\sim \beta_{10}+\beta_{11}\text{wind}+\beta_{12}\text{temperature}+\beta_{13}\text{temperature}\times \text{wind}
\end{aligned}
$$

## Computation

The fitting of the POMP models are based on the pomp package for R. Specifically I am using the iterated particle filtering algorithm to estimate the parameters. Because of the limitation of time, after I fit the third model and get the parameters of the measurement model, I set them as constant when fitting further models. I take this approach because I am not particularly interested in the parameters of the gamma distributions. This approach can also increase the efficiency of the algorithm given that there are fewer parameters that it needs to optimize. 

# Results

## Exploratory Data Analysis

First I plot the raw data and its autocorrelation for different lags. 
```{r}
eagle_data = read.csv("eagle_421.csv")
ggplot(eagle_data, aes(x=timestamp, y=msa))+geom_line()+
  xlab("Time Point")+ylab("Minimum Specific Acceleration(MSA)")
```

```{r}
acf(eagle_data$msa, main="MSA")
```

There is significant autocorrelation at small lags, which indicates that a POMP model could fit the data well.

## Log Likelihood Results

I summarize the log likelihoods of all the models in the table below. 

```{r}
simplegamma <- fitdist(eagle_data$msa, distr = "gamma", method = "mle")
model1lik <- simplegamma$loglik

load("result_nocovar_nomix.rda")
model2lik <- max(liks_global_nocovar_nomix)

load("result_nocovar_mix.rda")
model3lik <- max(liks_global_nocovar_mix)

load("global_search_covar1_mix-2.rda")
load("lik_global_eval_covar1_mix-2.rda")
model4lik <- max(liks_global_covar1_mix)

load("global_search_covar2_mix-2.rda")
load("lik_global_eval_covar2_mix-2.rda")
model5lik <- max(liks_global_covar2_mix)

load("global_search_covar3_mix-2.rda")
load("lik_global_eval_covar3_mix-2.rda")
model6lik <- max(liks_global_covar3_mix)

load("global_search_covar4_mix-2.rda")
load("lik_global_eval_covar4_mix-2.rda")
model7lik <- max(liks_global_covar4_mix)

myliks <- c(model1lik,model2lik,model3lik,model4lik,
            model5lik,model6lik,model7lik)
param_num <- c(2, 6, 9, 11, 13, 15, 17)
paperliks <- c(NA, NA, 2000.2, 2001.9, 2010.4, 2011.6, 2017.0)

resulttable <- data.frame(cbind(param_num,myliks,  paperliks))

colnames(resulttable) <- c("Number of Parameters", "Log Likelihood", "Results from Article")
rownames(resulttable) <- paste0("Model ",seq(1,7))
kable(resulttable, digits=2)
```

It's worth noting that the log likelihood from the article is calculated based on the data from all seven days, therefore it's much larger than my log likelihood calculated from a single day. After comparing the log likelihoods, I conclude that model 4 is the optimal model. Model 4 assumes that the transitional probability between hidden states can be fitted by a logistic regression with wind speed as the only covariate.

## Model Diagnostics

To evaluate the model diagnostics, I decide to offer the convergence diagnostics plots for model 3 and model 4. Recall that I used model 3 to decide the parameters for the measurement model and I used model 4 to decide the rest of the parameters for the transitional probability matrix.

First I offer the diagnostics plots for model 3, which is the POMP model without covariates.
```{r}
plot(mif_nocovar_mix)
```

All the iterative filtering trials converge, yet they converge to several distinct parameter sets. Given the time constraint, I took a heuristic and decide to pick the parameter sets based on two standards. The first standard is that the mean of the gamma distribution corresponding to my presumed active state should be higher than the mean values of the gamma distributions corresponding to my presumed inactive states. The second standard is high log likelihood. Luckily after I investigate all the filtering trials, the one with the highest log likelihood meets both my standards. Therefore I go ahead with this set of parameters.

Next is the diagnostics plot for model 4, which is the optimal model that I select.

```{r}
plot(mifs_global_covar1_mix)
```

All the iterative filtering trials converge to a single set of parameters, indicating that the model is stable.

We can also visually checkout the goodness of fit by simulating several time series using the selected model.

```{r}
source("selected_model.R")
selected_params = coef(mifs_global_covar1_mix)[,which.max(liks_global_covar1_mix)]

set.seed(10)

sims <- simulate(eagle_2, params=selected_params, 
                 nsim=4, format="data.frame", include=TRUE, t0=1)
ggplot(sims,mapping=aes(x=timestamp,y=msa,color=.id=="data"))+
  geom_line()+guides(color=FALSE)+facet_grid(rows=vars(.id))
```

Because of the unpredictability of the eagle's activity, the shape of my time series cannot match the eagle's exactly, but the overall trends are the same. Most of the time the acceleration is small, but occsasionally there are high peaks.

## Interpretation

The selected parameter is

```{r}
kable(selected_params)
```

This set of parameters tell us that
$$
\text{log odds}(p_{\text{inactive}\rightarrow \text{active}})=-2.87+0.14\times \text{wind}
$$

$$
\text{log odds}(p_{\text{active}\rightarrow \text{active}})=-1.73+1.32\times \text{wind}
$$

Within the active state, the MSA follows
$$
MSA\sim Gamma(\text{shape}=1.066, \text{scale}=0.356)
$$

Within the inactive state, the MSA follows a mixture of

$$
\begin{aligned}
MSA\sim Gamma(\text{shape}=20.51, \text{scale}=6.96\times 10^{-4})\\
MSA\sim Gamma(\text{shape}=2.972, \text{scale}=0.0091)
\end{aligned}
$$

It would be more illustrative with some plots. Inspired by the article, I plot the probability of the eagle staying in its original state against the wind speed.

```{r}
windspeed = seq(0, 6, 0.1)

stay0 = function(windspeed){
  loddsstay = -(selected_params["beta00"]+selected_params["beta01"]*windspeed)
  exp(loddsstay)/(1+exp(loddsstay))
}

stay1 = function(windspeed){
  loddsstay = (selected_params["beta10"]+selected_params["beta11"]*windspeed)
  exp(loddsstay)/(1+exp(loddsstay))
}

prob0 = stay0(windspeed)
prob1 = stay1(windspeed) 

tpm = data.frame(Probability = c(prob0, prob1),
                 State = rep(c("Active", "Inactive"), c(61,61)),
                 Windspeed = c(windspeed, windspeed))
```




# Discussion




